Demography and linked selection in wild and domesticated maize

### 2015-10-17 manuscript.tex
updates

\documentclass{pnastwo}
\usepackage{cite}
\usepackage{color}
\usepackage{pgf}
%\usepackage{hyperref}scho
\usepackage[normalem]{ulem}
\newcommand{\jri}[1]{\textcolor{red}{\scriptsize #1}}
\newcommand{\mbh}[1]{\textcolor{blue}{\scriptsize #1}}
\newcommand{\citex}{\textcolor{red}{\bf CITE}}
\newcommand{\X}{\textcolor{red}{\bf X}}
\usepackage{multibib}

@@

Our findings demonstrate that classic hard selective sweeps have not contributed substantially to genome-wide patterns of diversity in maize, a result we show is robust to concerns about power due to the effects of background selection  \cite{enard2014}. 
Although our approach ignores the potential for hard sweeps in noncoding regions of the genome, a growing body of evidence argues against hard sweeps as the prevalent mode of selection shaping maize variability. 
Among well-characterized domestication loci, only the gene \emph{tga1} shows evidence of a hard sweep on a missense mutation \cite{wang2015}, while several loci are consistent with ``soft sweeps''  from standing variation \cite{studer2011,gallavotti2004role} or multiple mutations \cite{wills2013}. 
Moreover, genome-wide studies of domestication \cite{hufford2012},  local adaptation \cite{Takuno15062015} and modern breeding \cite{beissinger2014} all support the importance of standing variation and/or non-coding variants as primary sources of adaptive variation. 
Soft sweeps are expected to be common when $2N_e\mu_b \ge 1$, where $\mu_b$ is the mutation rate of beneficial alleles with selection coefficient $s_b$ \cite{messer2013population}.  
Assuming a mutation rate of $3 \times 10^{-8}$ \cite{clark2005} and that on the order of $\approx 1-5\%$ of mutations are beneficial \cite{eyre2007distribution}, this implies that soft sweeps should be common in both maize and teosinte for mutational targets $>>10kb$ --- a plausible size for quantitative traits or for regulatory evolution targeting genes with large up- or down-stream control regions \cite[e.g.]{studer2011}.
Indeed, many adaptive traits in both maize \cite{wallace2014} and teosinte \cite{weber2008} are highly quantitative,  and adaptation in both maize \cite{hufford2012} and teosinte \cite{pyhajarvi2013complex} has involved selection on regulatory variation.
%\jri{what \% of the windows of top H12 score genome-wide occur outside of genes? is that a number to include here?}
% TB Says: No, because windows were 200 SNPs so even windows that don't correspond to a gene could include a gene.

@@

\subsection{Demography of domestication} %JRI
Although a number of authors have investigated the demography of maize domestication \cite{eyre1998, tenaillon2004selection, wright2005}, these  efforts relied on data only from genic regions of the genome and made a number of limiting assumptions about the demographic model.  We show that diversity within genes has been strongly reduced by the effects of linked selection, such that even synonymous polymorphisms in genes are not representative of diversity at unconstrained sites.  This implies that genic polymorphism data are unable to tell the complete or accurate demographic history of maize. Furthermore, by utilizing the full joint SFS, we are able to estimate population growth, gene flow, and the strength of the domestication bottleneck without requiring assumptions about its duration.

One surprising result from our model is the estimated timing of domestication at $\approx 15,000$ years before present.
While this appears to conflict with archaeological estimates \cite{piperno2009starch}, we instead argue this estimate reflects the fact that the genetic split between populations likely preceded anatomical changes that can be identified in the archaeological record. 
We also note that our result may also be inflated due to population structure, as our geographically diverse sample of teosinte may include populations diverged from those that gave rise to maize.

@@

\subsection{Demography influences the efficiency of purifying selection}
We find a more pronounced decrease in nucleotide diversity around genes in teosinte compared to domesticated maize (Figure \ref{fig:purify}A).
This difference is likely due to the effects of purifying selection.
Purifying selection is more efficient in larger populations  \cite{kimura1984}, and our observation is thus consistent with our demographic model and previous work \cite{eyre1998, tenaillon2004selection, wright2005, ross2009} showing a smaller long-term effective population size in maize.  
To understand how recent population growth has changed the effects of linked selection, we also analyzed diversity of singleton polymorphisms.   Under a simple neutral model, lower frequency polymorphisms are expected to be younger, and patterns of diversity at singleton SNPs should thus represent recent evolutionary history. 
Consistent with its larger current size and thus more efficient purifying selection, we see a stronger decline of singleton diversity around genes in maize (Figure \ref{fig:purify}B than teosinte.

A consequence of the inefficient purifying selection that maize experienced during its bottleneck is likely that it harbors more weakly deleterious alleles segregating at intermediate frequency than does teosinte. This could be a part of the explanation of why maize inbreds have continued to improve over the past several decades \cite{meghji1984}; if deleterious alleles tend to be recessive and are particularly frequent, they will have ample opportunities to display their phenotypes in inbreds. Our results also demonstrate that recent purifying selection in maize has become much more effective, potentially explaining the ongoing improvement of these inbreds as maize lines are continuously selected. Additionally, the large $N_e$ of modern maize compared to modern teosinte implies that for new mutations, selection will operate much more efficiently in maize.

The absence of evidence for a genome-wide impact of hard sweeps differs markedly from observations in \emph{Drosophila} \cite{sattath2011} and \emph{Capsella} \cite{williamson2014}, but is consistent with data from humans \cite{hernandez2011,pritchard2010genetics}. 
In a related study, \cite{corbett2015} showed that selection tends to have an elevated impact on reducing diversity in species with large population size compared to small. This may explain why maize and humans, both bottlenecked species, show similar patterns while drosophila and capsella behave differently. \textcolor{red}{However, we also do not observe abundant hard sweeps in teosinte, which calls this explanation into question.}

\jri{these results have implications for lots of studies. recent demographic history often ignored (examples). in crops, bneck is considered and loss of diversity, but expansion also likely ubiquitous and consequences not well studied.  in humans XYZ, and we see some of that already here. this argues that a full understanding of diversity in both wild and cultivated plants will require recent demography. mention C-D ignored recent demography -- here's where rec/pi windows could help.}

\jri{I would like to see a supplemental figure or maybe just number here on the partial correlation in both teosinte and maize of diversity with recombination after accounting for gene density.  do this in 500kb windows and then in 10kb windows.  we should do the correlation with both pi and singletons.  This let's us compare more explictly with studies like C-D who evaluate overall patterns of linked selection. will be especially cool if we don't see a big difference in big windows but we do in 10kb windows -- siggests C-D looked at wrong scale  }

\jri{inbreds stuff is OK, but should be shortened.  we are aiming at a broader audience that doesn't care about maize.  what are more/broader. Chia, Rogers-Melnick, Mezmouk all looked at load. does this have implications for large effect load (i.e. should bneck have purged those?), for heterosis, architecture of quant. traits, etc.}

>>>>

\documentclass{pnastwo}
\usepackage{cite}
\usepackage{color}
\usepackage{pgf}
%\usepackage{hyperref}scho
\usepackage[normalem]{ulem}
\newcommand{\jri}[1]{\textcolor{red}{\scriptsize #1}}
\newcommand{\mbh}[1]{\textcolor{blue}{\scriptsize #1}}
\newcommand{\tmb}[1]{\textcolor{brown}{\scriptsize #1}}
\newcommand{\citex}{\textcolor{red}{\bf CITE}}
\newcommand{\X}{\textcolor{red}{\bf X}}
\usepackage{multibib}

@@

Our findings demonstrate that classic hard selective sweeps have not contributed substantially to genome-wide patterns of diversity in maize, a result we show is robust to concerns about power due to the effects of background selection  \cite{enard2014}. 
Although our approach ignores the potential for hard sweeps in noncoding regions of the genome, a growing body of evidence argues against hard sweeps as the prevalent mode of selection shaping maize variability. 
Among well-characterized domestication loci, only the gene \emph{tga1} shows evidence of a hard sweep on a missense mutation \cite{wang2015}, while several loci are consistent with ``soft sweeps''  from standing variation \cite{studer2011,gallavotti2004role} or multiple mutations \cite{wills2013}. 
Moreover, genome-wide studies of domestication \cite{hufford2012},  local adaptation \cite{Takuno15062015} and modern breeding \cite{beissinger2014} all support the importance of standing variation and/or non-coding variants as primary sources of adaptive variation. 
Soft sweeps are expected to be common when $2N_e\mu_b \ge 1$, where $\mu_b$ is the mutation rate of beneficial alleles with selection coefficient $s_b$ \cite{messer2013population}.  \tmb{Jeff, you put this here, but as written I don't think we need to mention $s_b$}
Assuming a mutation rate of $3 \times 10^{-8}$ \cite{clark2005} and that on the order of $\approx 1-5\%$ of mutations are beneficial \cite{eyre2007distribution}, this implies that soft sweeps should be common in both maize and teosinte for mutational targets $>>10kb$ --- a plausible size for quantitative traits or for regulatory evolution targeting genes with large up- or down-stream control regions \cite[e.g.]{studer2011}.
Indeed, many adaptive traits in both maize \cite{wallace2014} and teosinte \cite{weber2008} are highly quantitative,  and adaptation in both maize \cite{hufford2012} and teosinte \cite{pyhajarvi2013complex} has involved selection on regulatory variation.

The absence of evidence for a genome-wide impact of hard sweeps differs markedly from observations in \emph{Drosophila} \cite{sattath2011} and \emph{Capsella} \cite{williamson2014}, but is consistent with data from humans \cite{hernandez2011,pritchard2010genetics}. \cite{corbett2015} recently showed that selection tends to have an elevated impact reducing diversity in species with large population size compared to small, which may explain why maize and humans, both bottlenecked species, show similar patterns while drosophila and capsella behave differently.

%\jri{what \% of the windows of top H12 score genome-wide occur outside of genes? is that a number to include here?}
% TB Says: No, because windows were 200 SNPs so even windows that don't correspond to a gene could include a gene.

@@

\subsection{Demography of domestication} %JRI
Although a number of authors have investigated the demography of maize domestication \cite{eyre1998, tenaillon2004selection, wright2005}, these  efforts relied on data only from genic regions of the genome and made a number of limiting assumptions about the demographic model.  We show that diversity within genes has been strongly reduced by the effects of linked selection, such that even synonymous polymorphisms in genes are not representative of diversity at unconstrained sites. This implies that genic polymorphism data are unable to tell the complete or accurate demographic history of maize, but the rapid recovery of diversity outside of genes demonstrates that sites far from genes can be reasonably used for demographic inference. Furthermore, by utilizing the full joint SFS, we are able to estimate population growth, gene flow, and the strength of the domestication bottleneck without requiring assumptions about its duration.

One surprising result from our model is the estimated timing of domestication at $\approx 15,000$ years before present.
While this appears to conflict with archaeological estimates \cite{piperno2009starch}, we emphasize that this estimate reflects the fact that the genetic split between populations likely preceded anatomical changes that can be identified in the archaeological record. 
We also note that our result may also be inflated due to population structure, as our geographically diverse sample of teosinte may include populations diverged from those that gave rise to maize.

@@

\subsection{Demography influences the efficiency of purifying selection}
One of our more striking findings is that the impact of purifying selection on maize and teosinte variability has varied through time. Although we observe a more pronounced decrease in $\pi$ around genes in teosinte than maize (Figure \ref{fig:purify}A), we see a greater decrease in singleton diversity in maize than teosinte (Figure \ref{fig:purify}B). Purifying selection is more efficient in larger populations  \cite{kimura1984}, and our observations using $\pi$  are thus consistent with our demographic model and previous work \cite{eyre1998, tenaillon2004selection, wright2005, ross2009} which shows a smaller long-term effective population size in maize. Similarly, our observations based on singleton diversity, which thereby reflect more recent patterns of selection, again are in agreement with maize and teosinte demography; from the viewpoint of a recent mutation, the rapid expansion of maize post-domestication means that today maize has a vastly larger effective population size than does teosinte.

A consequence of the previously-inefficient purifying selection in maize is that it likely harbors more weakly deleterious alleles segregating at intermediate frequency than does teosinte. Therefore, high genetic load could be a part of the explanation of why maize inbreds have continued to improve over the past several decades \cite{meghji1984}. Our results also demonstrate that recent purifying selection in maize has become much more effective, potentially explaining the ongoing improvement of these inbreds as maize lines are continuously selected. Additionally, the large $N_e$ of modern maize compared to modern teosinte implies that for new mutations, selection will operate much more efficiently in maize. This observation is not restricted to maize. Bottlenecks are have been documented or predicted across most domesticated plant and animal species \cite{doebley2006, hyten2006, zhu2007, bovine2009, larson2013 }, but rapid expansion post-bottleneck, which we demonstrate generates ubiquitous consequences relating to genetic diversity, is not well studied. In human, model, and other non-domesticated species, these results demonstrate that observed patterns of selection may depend heavily on recent and/or ancient patterns of demography.

%\jri{these results have implications for lots of studies. recent demographic history often ignored (examples). in crops, bneck is considered and loss of diversity, but expansion also likely ubiquitous and consequences not well studied.  in humans XYZ, and we see some of that already here. this argues that a full understanding of diversity in both wild and cultivated plants will require recent demography. mention C-D ignored recent demography -- here's where rec/pi windows could help.}

%\jri{I would like to see a supplemental figure or maybe just number here on the partial correlation in both teosinte and maize of diversity with recombination after accounting for gene density.  do this in 500kb windows and then in 10kb windows.  we should do the correlation with both pi and singletons.  This let's us compare more explictly with studies like C-D who evaluate overall patterns of linked selection. will be especially cool if we don't see a big difference in big windows but we do in 10kb windows -- siggests C-D looked at wrong scale  }
